The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-08 Papers/Universalis/ModalCollapse.tex
Add acknowledgment and note about authorship

Although G\"odel's axioms lead to modal collapse, 
there are several variants (e.g. \cite{anderson,bjordal}) 
that are known to be immune to the modal collapse. 
This means there must be at least one proposition $\phi$ 
such that $\phi \imp \nec \phi$ 
(from now on abbreviated $\collapse{\phi}$) is not valid 
under the axioms and definitions used by the variant. 
But if the variant is sufficiently similar to G\"odel's argument, 
following the path deriving Lemmas \textbf{L1} and 
\textbf{L2}, then 
$\collapse{\exq x G(x)}$ must be valid. 
Therefore, one may wonder how strong is their immunity to 
the modal collapse: is there any other proposition $\phi$ 
for which $\collapse{\phi}$ is also valid?

For Anderson's emendation \cite{anderson}, for example, a form of the modal collapse (\textbf{A:MC}), restricted to positive properties applied to god-like beings, can be derived. The proof, under the modal logic \textbf{\emph{K}}, depends only on Anderson's alternative definition of god-like being (\textbf{A:D1}). This class of propositions for which the collapse occurs is tight: weaker restrictions (\textbf{A:MC1} and \textbf{A:MC2}), which could lead to larger classes, are counter-satisfiable.

@@

The technique enabling these experiments is the embedding of quantified modal logics into higher-order logics \cite{J23,B9,C36}, for which automated theorem provers exist. This technique has already been succesfully employed in the verification and reconstruction of G\"odel's proof \cite{todo:all our previous papers}, and a detailed mathematical description is available in \cite{todo:ECAI}.

The modal collapse is an interesting example of philosophical controversy and dispute, to which Leibniz's idea of a \emph{calculus ratiocinator} in the form of contemporary automated theorem provers can be applied. Current technology is increasingly ready to be embraced by those willing to practice computer-assisted theoretical philosophy \cite{oppenheimera11,rushby13}.

@@

% ToDo: Leibniz calculemus, Rushby, Zalta

% \begin{defn}
% This serves as environment for definitions. Note that the text
% appears not in italics.
% \end{defn}

% \begin{equation}\label{testequation}
% \text{This is a sample equation: } c^2=a^2+b^2
% \end{equation}

% \begin{thm}[Main Theorem]
% In contrast to definitions, theorems appear typeset in italics as
% it has become more or less standard in most textbooks and
% monographs. Equations can be cited using the \verb+\eqref+ command which
% automatically adds brackets: \verb+\eqref{testequation}+ results in \eqref{testequation}.
% \end{thm}

% \begin{proof}
% A special environment is predefined: the \textit{proof} environment. Please use
% \begin{verbatim}\begin{proof}\end{verbatim}
% proof of the statement
% \begin{verbatim}\end{proof}\end{verbatim}
% for typesetting your proofs. The end-of-proof symbol $\Box$ will be added automatically.
% \end{proof}

% There are two known problems with the placement of the end-of-proof sign:

% \begin{enumerate}
%   \item if your proof ends with a\ \ s i n g l e\ \ displayed line, the end-of-proof sign would
% be placed in the line below; if you want to avoid this, write your line in the form
% \begin{verbatim}$$displayed math line \eqno\qedhere$$\end{verbatim}
% which results in

% \begin{proof}
% $$displayed math line \eqno\qedhere$$
% \end{proof}
% \item if your proof ends with an aligned displayed environment, the command
% \verb+\tag*{\qed}+ can be used to place the end-of-proof sign properly:
% \begin{verbatim}
% \begin{align*}
% \alpha&=\beta+\gamma\\
% &=\delta+\epsilon\tag*{\qed}
% \end{align*}
% \end{verbatim}
% results in
% \begin{align*}
% \alpha&=\beta+\gamma\\
% &=\delta+\epsilon\tag*{\qed}
% \end{align*}
% \end{enumerate}
% Please try to avoid using the obsolete \verb+\eqnarray+ environment. This environment has several bugs
% and has been replaced by the more flexible \AmS\ environments \verb+align, split, multline+.

% \begin{rem}
% Additional comments are being typeset without boldfaced entrance
% word as they may be minor important.
% \end{rem}

% \begin{ex}
% For some constructs, even no number is required.
% \end{ex}

% Displayed equations may be numbered like the following one:
% \begin{equation}
% \sqrt{1-\sin^2(x)}=|\cos(x)|.
% \end{equation}

% ------------------------------------------------------------------------

% \subsection*{Acknowledgment}
% Many thanks to ...

>>>>

Although G\"odel's axioms lead to modal collapse, 
there are several variants (e.g. \cite{anderson,bjordal}) 
that are known to be immune to the modal collapse. 
This means there must be at least one proposition $\phi$ 
such that $\phi \imp \nec \phi$ 
(from now on abbreviated as $\collapse{\phi}$) is not valid 
under the axioms and definitions used by the variant. 
But if the variant is sufficiently similar to G\"odel's argument, 
following the path deriving Lemmas \textbf{L1} and 
\textbf{L2}, then 
$\collapse{\exq x G(x)}$ must be valid. 
Therefore, one may wonder how strong is their immunity to 
the modal collapse: is there any other proposition $\phi$ 
for which $\collapse{\phi}$ is also valid?

For Anderson's emendation \cite{anderson}, for example, a form of the modal collapse (\textbf{A:MC}), restricted to positive properties applied to god-like beings, can be derived. The proof, under the modal logic \textbf{\emph{K}}, depends only on Anderson's alternative definition of god-like being (\textbf{A:D1}). This class of propositions for which the collapse occurs is tight: weaker restrictions (\textbf{A:MC1} and \textbf{A:MC2}), which could lead to larger classes, are counter-satisfiable. These results hold under both constant and varying domain quantification, with possibilist and actualist quantifiers.

@@

The technique enabling these experiments is the embedding of 
quantified modal logics into higher-order logics 
\cite{J23,B9,C36}, for which automated theorem provers exist. 
This technique has already been successfully employed in the 
verification and reconstruction of G\"odel's proof 
\cite{todo:all our previous papers}, and a detailed 
mathematical description is available in \cite{todo:ECAI}.

The modal collapse is an interesting example of philosophical 
controversy and dispute, to which we can apply Leibniz's idea 
of a \emph{calculus ratiocinator} brought to reality in the 
form of contemporary automated theorem provers. 
A significant advantage provided by the use of computers is 
that all conditions (e.g. modal logic, domain conditions, 
semantics) under which the announced results hold must be 
explicitly written in the source code. This reduces the need 
for interpretation and the danger of misunderstandings. 
Current technology is increasingly ready to be embraced by 
those willing to practice computer-assisted theoretical 
philosophy \cite{oppenheimera11,rushby13}.

@@

\subsection*{Acknowledgment}

We would like to thank Paul Weingartner, Andr\'{e} Fuhrmann and Melvin Fitting for several discussions about G\"odel's ontological argument.

ToDo: who else?

\subsection*{Note about authorship}

Alphabetic order has been used for the authors' names. The extent and kind of contribution of each author cannot be inferred from the order.
